
---
title: "Exploring Penguin Bill Measurements"
author: "Rakshit Kapoor"
date: 2026-09-24
categories: [R, Data Analysis, Visualization]
---

## Overview

In this post, I investigate how bill length and bill depth differ among Adelie, Chinstrap, and Gentoo penguins. Using R, I calculate summary statistics and create visualizations to explore differences between the three species.

The data come from the [Palmer Penguins dataset](https://allisonhorst.github.io/palmerpenguins/), collected by Dr. Kristen Gorman and the Palmer Station Antarctica LTER.

## 1. Load and explore the data

First, I load the required R packages and inspect the dataset.

```{r}
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(ggplot2)

penguins_data <- palmerpenguins::penguins

head(penguins_data)
```

The dataset contains measurements of penguins from three species, including bill length, bill depth, flipper length, and body mass.

## 2. Calculate summary statistics

Next, I calculate the average bill length and bill depth for each species. I exclude missing measurements to avoid affecting the calculations.

```{r}
bill_summary <- penguins_data |>
  group_by(species) |>
  summarise(
    avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
    avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

bill_summary
```

These statistics allow us to compare the typical bill measurements across the three species.

## 3. Visualize bill measurements

I create a scatter plot to investigate the relationship between bill length and bill depth. Each species is represented by a different colour.

```{r}
#| label: fig-penguin-bills
#| fig-cap: "Relationship between bill length and bill depth across three penguin species."
#| warning: false

penguins_data |>
  ggplot(aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    colour = species
  )) +
  geom_point(alpha = 0.7, na.rm = TRUE) +
  labs(
    title = "Penguin Bill Length vs Bill Depth",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    colour = "Species"
  ) +
  theme_minimal()
```

The visualization helps identify differences in bill measurements between the species. It also shows how bill length and depth vary within each species.

## 4. Compare average bill length

Finally, I arrange the species by their average bill length to identify differences between them.

```{r}
bill_ranking <- bill_summary |>
  arrange(desc(avg_bill_length))

bill_ranking
```

The resulting table orders the three species by average bill length, making it easier to compare their measurements.

## Conclusion

This analysis demonstrates how R can be used to summarize and visualize biological measurements.

The summary statistics and scatter plot reveal differences in bill morphology among the three penguin species. These differences illustrate how grouping, aggregation, and visualization can help explore patterns in real-world datasets.